Human judgement forecasting of COVID-19 in the UK

### 2024-02-14 manuscript/manuscript.tex
Update manuscript to incorporate feedback from Anne Cori

This paper represents a follow-up study to \citet{bosseComparingHumanModelbased2022} in the United Kingdom with forecasts made over the course of thirteen weeks between May and August 2021. Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application. All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022}. This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country, in a different time period, and with an increased number of participants. In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the estimated effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a simulation model. We describe this approach as human in the loop computational modelling and consider it a formalisation of often practiced manual intervention in computational forecasts.

@@

The study period (forecasts were made between May 24 and August 16, 2021, for targets between May 29 and September 11, 2021) was characterised by an increase in the number of cases and deaths in the United Kingdom. Reported cases in particular rose rapidly compared to pre-study levels, with a peak on July 17, 2021, followed by a trough and another subsequent increase in numbers. Death numbers remained almost constant in the first four weeks of the study period, followed by a steady increase until the end of the study period in September 2021. This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021} as well as the  European Football Championship \citep{dehningImpactEuro20202023}. An overview of the reported case and death numbers is shown in Figure \ref{fig:overview-data}.

\begin{figure*}
\centering
\includegraphics[width=0.99\textwidth]{../output/figures/overview-data.png}
\caption{\bf{Observed cases and deaths of COVID-19 in the UK}. A: Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021. Daily numbers were multiplied by seven in order to appear on the same scale as weekly numbers.}
\label{fig:overview-data}
\end{figure*}

@@

% Study period. 
The study period was marked by an increase in both case and death numbers. Case numbers rose quickly compared to the pre-study period, peaking on July 17, 2021, followed by a trough and a subsequent further increase. Forecasts displayed a familiar pattern, where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak. Death numbers were increasing more slowly than during the previous peak in January 2021, in line with the beginning of vaccination efforts and a greater natural immunity in the population. The peak in case numbers did not correspond to a peak in death numbers, suggesting some decoupling of case and death numbers such as would be expected from effects of immunity that are stronger in preventing severe disease than any symptoms. Forecasters tended to overpredict death numbers in the beginning, while underpredicting them in the end, expecting death numbers to fall after the peak in cases. The study period coincides with the rise of the Delta variant in the UK \citep{bastIncreasedRiskHospitalisation2021}, as well as the 2021 European Football Championship, which likely shifted the age distribution towards younger cases \citep{dehningImpactEuro20202023}.

@@

In their original study, conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths. They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators. \citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19. 
Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it. In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, and worse on the natural scale. Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths. 
During the study period, the case fatality ratio (CFR) likely changed quite quickly compared to the pre-study period. On the one hand, the rise of the Delta variant in the UK, which was first discovered in the UK in March 2021, was estimated to have a higher CFR than previous variants \citep{bastIncreasedRiskHospitalisation2021, twohigHospitalAdmissionEmergency2022}. On the other hand, the ongoing COVID-19 vaccination and growing natural immunity in the population had decreasing effects on the CFR. In addition, the age distribution of cases changed throughout the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}. Overall, the CFR was lower than during previous peaks of COVID-19.
One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update. The present study only saw a steady increase in death numbers, which one could argue is relatively easy to predict, making it difficult to compare forecast performance with performance in other settings. A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.

Importantly, in this study our combined crowd ensemble ("epiforecasts-EpiExpert") contributed to the European Forecast Hub ensemble. This in contrast to the study by \citet{bosseComparingHumanModelbased2022}, where they compared crowd forecasts against a hypothetical ensemble excluding the crowd forecasts. In the original study, including the crowd forecasts improved the Hub ensemble on average (however, the overall number of models included in the German and Polish Hub ensemble was smaller than the number of models in the European Forecast Hub ensemble). In our study, comparisons between our crowd ensembles and the Forecast Hub ensemble are therefore confounded by the fact the combined crowd ensemble was included in the Forecast Hub ensemble, possibly leading us to underestimate differences between the two.

@@

This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths. One appeal of this approach is that the forecaster can directly forecast the generative process and how they believe it is affected by interventions and changes in behaviour. Computational modelling then takes care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, and the relationship between different indicators. This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$. Though we note all of these modelling steps and the construction of the model itself requires the human constructing the model to make assumptions. Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience. In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more. Sample sizes, however, were very low. 
Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
$R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths. The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon. However, as mentioned before, the CFR likely evolved during the study period. Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast, likely impacting forecast accuracy. They also had no way to adjust the CFR manually. Giving human forecasters to see their implied death forecasts, as well as the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model. Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period. Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters. Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely negatively affected forecasts. One potential way to disentangle the effect of the convolution model from the $R_t$ forecasts would have been to use the human forecasts for cases as an input to the second computational model, which could then have simulated deaths. Future work could expose forecasters to different combinations of these options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.

>>>>

This paper represents a follow-up study to \citet{bosseComparingHumanModelbased2022} in the United Kingdom with one- to four-week ahead forecasts made over the course of thirteen weeks between May 24 and August 16, 2021. The study period is after the second wave of COVID-19 in the UK (which peaked in January 2021) and falls into a time when restrictions in the UK were gradually lifted as part of the roadmap out of lockdown (with final restrictions lifted on July 19, 2021). The forecast period also coincides with the rise of the Delta variant \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023} starting April 2021.

Forecasts were elicited from experts and laypeople as part of a public forecasting tournament, the "UK Crowd Forecasting Challenge" using a web application. All forecasts were submitted to the European COVID-19 Forecast Hub, one of several Forecast Hubs that have been systematically collating forecasts of different COVID-19 forecast targets in the US \citep{cramerEvaluationIndividualEnsemble2021}, Germany and Poland \citep{bracherShorttermForecastingCOVID192021, bracherNationalSubnationalShortterm2022}, and Europe \citep{sherrattPredictivePerformanceMultimodel2022}. This study aims to investigate whether the original findings in \citet{bosseComparingHumanModelbased2022} with respect to forecaster performance replicate in a different country, in a different time period, and with an increased number of participants. In addition, it explores the approach proposed in \citet{bosseComparingHumanModelbased2022} to ask participants for a forecast of the estimated effective reproduction number $R_t$ which is then translated into a forecast of cases and deaths using a simulation model. We describe this approach as human in the loop computational modelling and consider it a formalisation of often practiced manual intervention in computational forecasts.

@@

The study period (forecasts were made between May 24 and August 16, 2021, for targets between May 29 and September 11, 2021) was characterised by an increase in the number of cases and deaths in the United Kingdom. Reported cases in particular rose rapidly compared to pre-study levels, with a peak on July 17, 2021, followed by a trough and another subsequent increase in numbers. Death numbers remained almost constant in the first four weeks of the study period, followed by a steady increase until the end of the study period in September 2021. This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023} as well as the  European Football Championship \citep{dehningImpactEuro20202023}. An overview of the reported case and death numbers is shown in Figure \ref{fig:overview-data}.

\begin{figure*}
\centering
\includegraphics[width=0.99\textwidth]{../output/figures/overview-data.png}
\caption{\bf{Observed cases and deaths of COVID-19 in the UK}. Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021. The green rectangle marks the study period from May 24 until September 11, 2021. Daily numbers were multiplied by seven in order to appear on the same scale as weekly numbers.}
\label{fig:overview-data}
\end{figure*}

@@

% Study period. 
The study period was marked by an increase in both case and death numbers. Case numbers rose quickly compared to the pre-study period, peaking on July 17, 2021, followed by a trough and a subsequent further increase. Forecasts displayed a pattern where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak. Similar patterns have been observed previously in other short-term forecasts of COVID-19 (see e.g, \citep{sherrattPredictivePerformanceMultimodel2022, bracherShorttermForecastingCOVID192021, bosseComparingHumanModelbased2022}).

Death numbers during the study period were increasing more slowly than during the previous peak in January 2021, coinciding with the beginning of vaccination efforts and a growing immunity in the population \citep{perez-guzmanEpidemiologicalDriversTransmissibility2023}. The peak in case numbers in July 2021 was not followed by a subsequent peak in death numbers (but rather a steady incline over several months), suggesting some decoupling of case and death numbers such as would be expected from effects of immunity that are stronger in preventing severe disease than any symptoms. Forecasters tended to overpredict death numbers in the beginning, while underpredicting them in the end, expecting death numbers to fall after the peak in cases. The study period coincides with the rise of the Delta variant in the UK \citep{bastIncreasedRiskHospitalisation2021, perez-guzmanEpidemiologicalDriversTransmissibility2023}, as well as the 2021 European Football Championship, which likely shifted the age distribution towards younger cases \citep{dehningImpactEuro20202023}.

@@

In their original study, conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths. They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators. \citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19. 
Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it. In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, and worse on the natural scale. Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths. 
During the study period, the case fatality ratio (CFR) likely changed quite quickly compared to the pre-study period. On the one hand, the rise of the Delta variant in the UK, which was first detected in the UK in March 2021, was estimated to have a higher CFR than previous variants \citep{bastIncreasedRiskHospitalisation2021, twohigHospitalAdmissionEmergency2022} (although \citep{perez-guzmanEpidemiologicalDriversTransmissibility2023} estimated it to be lower than that of the Alpha variant). On the other hand, the ongoing COVID-19 vaccination and growing natural immunity in the population had decreasing effects on the CFR. In addition, the age distribution of cases changed (hence modifying the overall CFR) throughout the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}. Overall, the CFR was lower than during previous peaks of COVID-19 \citep{perez-guzmanEpidemiologicalDriversTransmissibility2023}.
One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update. The present study only saw a steady increase in death numbers, which one could argue is relatively easy to predict, making it difficult to compare forecast performance with performance in other settings. A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.

Importantly, in this study our combined crowd ensemble ("epiforecasts-EpiExpert") contributed to the European Forecast Hub ensemble. This is in contrast to the study by \citet{bosseComparingHumanModelbased2022}, where they compared crowd forecasts against a hypothetical ensemble excluding the crowd forecasts. In the original study, including the crowd forecasts improved the Hub ensemble on average (however, the overall number of models included in the German and Polish Hub ensemble was smaller than the number of models in the European Forecast Hub ensemble). In our study, comparisons between our crowd ensembles and the Forecast Hub ensemble are therefore confounded by the fact the combined crowd ensemble was included in the Forecast Hub ensemble, possibly leading us to underestimate differences between the two.

@@

This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths. One appeal of this approach is that the forecaster can directly forecast the generative process and how they believe it is affected by interventions and changes in behaviour. Computational modelling then takes care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, and the relationship between different indicators. This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$. Though we note all of these modelling steps and the construction of the model itself requires the human constructing the model to make assumptions. Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience. In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more. Sample sizes, however, were very low. 
Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
$R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths. The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon. However, as mentioned before, the CFR likely evolved during the study period. Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast, likely impacting predictive performance. They also had no way to adjust the CFR manually. Giving human forecasters to see their implied death forecasts, as well as the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model. Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period. Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters. Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely negatively affected forecasts. One potential way to disentangle the effect of the convolution model from the $R_t$ forecasts would have been to use the human forecasts for cases as an input to the second computational model, which could then have simulated deaths. Future work could expose forecasters to different combinations of these options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.
